The relative roles of genetic, ecological and environmental factors in insular species radiations

### 2018-07-05 paper/draft.tex
more edits to intro, include refs

%% Cite & look at Katja Schiffer paper
%% "species rich vs. species poor lineage''
%% ``emerging systematic, functional, phylogenetic patterns''
%% Hurlbert
% !! Hypothesis: Quick establishment of reproductive barriers (+ population variability) facilitates/speeds up adaptation. !!
% Rationale: If genetic factors have impact on systems with clearly defined isolation (islands),
% they should be even more important on continental systems.

@@

Island radiations like the classical Darwin finches on Galápagos or \textit{Aeonium} on the Canaries have fascinated researchers throughout history. %% Maybe other examples?
But why were these genera able to produce this multitude of species as opposed to most other genera that remain species poor?

@@

(b) have reproductive barriers (e.g. environmental/geographical, ecological or behavioral) that facilitate establishment of distinct distribution patterns of said characteristic.

@@

Genetic analyses have already helped clearing up phylogenies of many species groups \cite{valenteUsingMolecularPhylogenies2018}%, like bats or cichlids REFS!

@@

Some genetic studies highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}

@@

In an island context, genetic studies are even rarer and almost exclusively of a descriptive nature.

%% Models
The problem of investigating speciation processes is the time frame and study extent needed for a proper evaluation.
Because island biogeography processes and thus radiations generally take place at evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult and virtually impossible to obtain.
Therefore, there has been an interest in building process-based island models to test the causality of assumptions and hypotheses.
In a recent review, we indeed found a rapid increase in number of mechanistic simulation models for island biogeography \cite{leidingerBiodiversityDynamicsIslands2017}.
Although many of the reviewed models integrate multiple processes at a time, unfortunately,
most models failed to include micro-evolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017).
One direction towards this is given by models using concepts such as ``protracted speciaton'' \cite{rosindellUnifiedModelSpecies2013}.
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this (cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} but see \cite{herbenGhostHybridizationNiche2005}) 
and genetic mechanisms should also play a central role for speciation/radiation and species richness.
Questions, such as "What functional traits (e.g. relating to dispersal capacity,
reproduction, trophic ecology) are associated with high diversification rates within and across island systems?"
or "What is the influence of gene flow among islands and/or
between islands and mainland areas on speciation rates?", thus remain largely open \cite{patinoRoadmapIslandBiology2017}
and may require explicit consideration of evolutionary processes.
%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
%%, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the most mechanisms integrated in island models [?] so far include only ecological
or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
while genetic properties are hardly recognized in the island context.
Given the essential role islands have played in the development of evolutionary theories, it is surprising that these processes
are not implemented more often in island simulations.

>>>>

%% Q24. What is the relative importance of ecological versus geographical speciation on islands?
%% => Patino hardly considering genetic mechanisms
%% ``emerging systematic, functional, phylogenetic patterns''
%% Hurlbert

% Rationale: If genetic factors have impact on systems with clearly defined isolation (islands),
% they should be even more important on continental systems.

@@

To this day, island radiations like the classical Darwin finches or Galápagos tortoises %%or \textit{Aeonium} on the Canaries
still engage researchers throughout the world \cite{roman-palaciosTortoiseFinchTesting}.
But why were these lineages able to produce this multitude of species as opposed to most other lineages that remain species poor?

@@

(b) have gene-flow restricting factors (e.g. environmental/geographical, ecological or behavioral) that facilitate establishment of distinct distribution patterns of said characteristic.

@@

Genetic analyses have already helped clearing up phylogenies and island biogeography of many species groups \cite{austinReconstructingIslandRadiation2004}\cite{emersonDiversificationForestBeetle2005}\cite{valenteUsingMolecularPhylogenies2018}%, like bats or cichlids REFS!

@@

Some genetic studies from continental species highlight the importance of genetic linkage or ``hitchhiking'' on reproductive/ecological isolation of populations \cite{hawthorneGeneticLinkageEcological2001}\cite{viaGeneticMosaicSuggests2008}

@@

Thus, genetic characteristics presents a powerful opportunity for a deeper understanding of speciation processes.

%% Models
However, the problem of investigating speciation processes is the time frame and study extent needed to go beyond mere correlation \cite{dormannCorrelationProcessSpecies}.
Because island biogeography processes and thus radiations generally take place at evolutionary time frames (millions of years),
comprehensive and meaningful data is difficult to obtain (\cite{didierLikelihoodTreeTopologies2017}, but see \cite{mitchellInferringDiversificationRate})
Therefore, there has been an interest in building process-based island models to test the causality of assumptions and hypotheses.
In a recent review, we indeed found a rapid increase in number of mechanistic simulation models for island biogeography \cite{leidingerBiodiversityDynamicsIslands2017}.
Although many of the reviewed models integrate multiple processes at a time, unfortunately,
most models failed to include micro-evolutionary or more explicit genetic processes %\cite{leidinger2017biodiversity}
(Leidinger \& Sarmento Cabral 2017).
One direction towards this is given by models using concepts such as ``protracted speciaton'' \cite{rosindellUnifiedModelSpecies2013}.
where speciation occurs with a delay from the actual
divergence event if the population remains isolated during that time.
But nature is usually more complex than this (cf.\ \cite{silvertownGhostCompetitionPhylogeny2004} but see \cite{herbenGhostHybridizationNiche2005}) 
and we expect genetic mechanisms to play a central role in speciation/radiation and species richness.
Questions, such as "What functional traits (e.g. relating to dispersal capacity,
reproduction, trophic ecology) are associated with high diversification rates within and across island systems?" %better another...
or "What is the influence of gene flow among islands and/or
between islands and mainland areas on speciation rates?", thus remain largely open \cite{patinoRoadmapIslandBiology2017}
and may require explicit consideration of evolutionary processes.
%% Yet, there are numerous examples for fascinating patterns of trait distributions and evolution on islands, such as island syndromes[, ...] or the fact of explosively radiating lineages (Hughes \& Eastwood 2006, Kocher 2004)
%%, while others remain one single species (Price/Chase) (eg. \cite{price2004speciation}).
%% Also: Ecological Opportunity and Adaptive Radiation, Stroud \& Losos 2016: ``Does the concept have predictive value[...]?'' \cite{stroud2016ecological}
%% Phylogenetic niche conservatism, phylogenetic signal and the relationship between phylogenetic relatedness and ecological similarity among species, Losos 2008: ecological vs. phylogenetic similarity \cite{losos2008phylogenetic}
Still, the most mechanisms integrated in island models [?] so far include only ecological
or geo-morphological processes (Rundell \& Price 2009, Esselstyn et al. 2011),
while genetic properties are hardly recognized in the island context.
Given the essential role islands have played in the development of evolutionary theories, it is surprising that these processes
are not implemented more often in island simulations.
